---
title: "Modelling"
author: "Taavi Päll"
date: "17 10 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Estonian Land Board data

Available via [www.maaamet.ee](http://www.maaamet.ee/kinnisvara/htraru/FilterUI.aspx).
API can be accessed programmatically using html requests.
One can use "rvest" package. Please have a look at "R/maaamet.R" script in [rstats-tartu/datasets](https://www.github.com/rstats-tartu/datasets) repo.
Returns html, but that ok too.
Html results need little wrangling to get table.

Data that can be downloaded are number of transactions and price summaries per property size, and type for different time periods.
Price info is given for data splits with more than 5 transactions.

## Load dataset

```{r}
##  Check if file is alredy downloaded 
if(!file.exists("data/transactions_residential_apartments.csv")){
  url <- "https://raw.githubusercontent.com/rstats-tartu/datasets/master/transactions_residential_apartments.csv"
  
  ## Check if data folder is present
  if(!dir.exists("data")){
    dir.create("data")
  }
  ## Download file to data folder
  download.file(url, "data/transactions_residential_apartments.csv")
}
```

## Import
```{r}
library(tidyverse)
library(lubridate)
library(stringr)
# Please try broom from GitHub as CRAN version may give obscure warning
# with augment
# devtools::install_github("dgrtwo/broom")
library(broom)
library(viridis)
apartments <- read_csv("data/transactions_residential_apartments.csv")
apartments <- apartments %>% 
  mutate(date = ymd(str_c(year, month, 1, sep = "-"))) %>% 
  select(date, everything())
harju <- apartments %>% filter(str_detect(county, "Harju"))
harju
```

## Strategy

- Start with single unit and identify interesting pattern

- Summarise pattern with model

- Apply model to all units

- Look for units that don't fit pattern

- Summarise with single model


## First glimpse

Plot **number of transactions** per year for Harju county and add smooth line.
```{r}
p <- ggplot(harju, aes(factor(year), transactions, group = area, color = area)) +
  geom_point() +
  geom_smooth(method = 'loess') +
  facet_wrap(~county, scales = "free_y") +
  labs(
    y = "Transactions",
    x = "Year"
  ) +
  scale_color_viridis(discrete = TRUE)
p
```

Mean price per unit area:
```{r}
## Add date 
p <- harju %>%
  ggplot(aes(date, price_unit_area_mean, group = area, color = area)) +
  geom_line() +
  geom_vline(xintercept = ymd("2008-09-15"), linetype = 2) +
  labs(title = "Transactions with residential apartments",
       subtitle = "Harju county",
       x = "Date",
       y = bquote(Mean~price~(eur/m^2)),
       caption = str_c("Source: Estonian Land Board, transactions database.\nDashed line, the collapse of the investment bank\nLehman Brothers on Sep 15, 2008.")) +
  scale_color_viridis(discrete = TRUE, name = bquote(Apartment~size~m^2))
p
```

Adjust mean price to changes in consumer index:
```{r}
download.file("https://raw.githubusercontent.com/rstats-tartu/datasets/master/consumer_index.csv",
              "data/consumer_index.csv")
consumer_index <- read_csv("data/consumer_index.csv")
consumer_index
## check if 2005 is 100%
consumer_index[10, 3:14] %>% rowMeans()
divide_by_100 <- function(x) {
  x/100
  }
consumer_index <- consumer_index %>% 
  select(-X1) %>% 
  gather(key = month, value = consumerindex, -year) %>% 
  mutate_at("consumerindex", divide_by_100)
consumer_index
```

Join apartments and consumer index data:
```{r}
apartments <- left_join(apartments, consumer_index, by = c("year", "month"))
```

Select harju county data:
```{r}
harju_ci <- filter(apartments, str_detect(county, "Harju"))
harju_ci
harju_ci <- harju_ci %>% 
  mutate(pua_mean_ci = price_unit_area_mean / consumerindex)
```


Price per m^2 after consumer index adjustment:
```{r}
harju_ci %>%
  ggplot(aes(date, pua_mean_ci, group = area, color = area)) +
  geom_line() +
  geom_vline(xintercept = ymd("2008-09-15"), linetype = 2) +
  labs(title = "Transactions with residential apartments",
       subtitle = "Harju county",
       x = "Date",
       y = "Consumer index corrected mean price\nrelative to 2005 (eur/m^2)",
       caption = str_c("Source: Estonian Land Board, transactions database.\nDashed line, the collapse of the investment bank\nLehman Brothers on Sep 15, 2008.")) +
  scale_color_viridis(discrete = TRUE, name = bquote(Apartment~size~m^2))
```

## Seasonal pattern

Seasonal pattern in number of transactions? Mean price eur/m^2 per month:
```{r}
## Note that we are rearranging x axis using R builtin vector of month names
p <- harju %>%
  ggplot(aes(factor(month, levels = month.abb), transactions, group = year)) +
  geom_line(alpha = 0.3) +
  geom_vline(xintercept = ymd("2008-09-15"), linetype = 2) +
  facet_wrap(~ area, scales = "free_y") +
  labs(title = "Seasonal trend in number of transactions\nwith residential apartments",
       subtitle = "Harju county",
       x = "Month",
       y = "Transactions",
       caption = "Source: Estonian Land Board, transactions database.") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
p
```

Add trendline per month. 
Fit linear model per month for each apartment size class.
`augment()` function from "broom" library adds fitted values and residuals to the original data.
We can use this to overlay model fit to our data.
```{r}
predicted_transa <- harju %>% 
  lm(transactions ~ month + area, data = .) %>%
  augment() # broom
predicted_transa %>% 
  arrange(desc(.cooksd)) %>% 
  dplyr::select(.cooksd, everything()) %>% 
  head
```

Overlay fitted values to previous plot.
```{r}
p + geom_line(data = predicted_transa, aes(month, .fitted, group = 1), 
              color = "red",
              size = 2) +
  labs(caption = "Red line: model fit of seasonal effect to number of transactions.\nSource: Estonian Land Board, transactions database.")
```
The number of transactions increases during period from March to May?


Do we have similar effect to average price per m^2?
```{r}
predicted_pua <- harju %>% 
  lm(price_unit_area_mean ~ month + area, data = .) %>%
  augment()
predicted_pua %>% head
```

```{r}
harju %>%
  ggplot(aes(factor(month, levels = month.abb), price_unit_area_mean, group = year)) +
  geom_line(alpha = 0.3) +
  geom_vline(xintercept = ymd("2008-09-15"), linetype = 2) +
  facet_wrap(~ area, scales = "free_y") +
  labs(title = bquote(Seasonal~trend~price~per~m^2~of~residential~apartments),
       subtitle = "Harju county",
       x = "Month",
       y = bquote(Mean~price~per~unit~area~(eur/m^2)),
       caption = "Source: Estonian Land Board, transactions database.") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  geom_line(data = predicted_pua, aes(month, .fitted, group = 1), 
              color = "red", 
              size = 2)
```



Residuals? Add also year to model to visualise residuals.
```{r}
predicted_trans2 <- harju %>% 
  lm(transactions ~ month + year + area, data = .) %>% 
  augment
predicted_trans2 %>% 
  head()
predicted_trans2 %>% 
  ggplot(aes(factor(month, levels = month.abb), .resid, group = year)) + 
  geom_hline(yintercept = 0, colour = "white", size = 3) +
  geom_line() + 
  facet_wrap(~ area)
```

For some apartement size classes, few years may have large influence on average number of transactions. We can look at the large cook's distance.
```{r}
predicted_trans2 %>% 
  arrange(desc(.cooksd)) %>% 
  dplyr::select(.cooksd, everything()) %>% 
  head()
```

Same for mean price per m^2:
```{r}
predicted_pua2 <- harju %>% 
  lm(price_unit_area_mean ~ month + year + area, data = .) %>% 
  augment 
predicted_pua2 %>% 
  ggplot(aes(factor(month, levels = month.abb), .resid, group = year, color = factor(year))) + 
  geom_hline(yintercept = 0, colour = "white", size = 3) +
  geom_line() + 
  facet_wrap(~ area) +
  scale_color_viridis(discrete = TRUE) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

Seems ok! 
```{r}
predicted_pua2 %>%
  arrange(desc(.cooksd)) %>% 
  dplyr::select(.cooksd, everything(.)) %>% 
  head()
```



Robust lm fit to number of transactions. Play with weights: price_min, price_unit_area_mean etc..
```{r}
# install.packages("MASS")
library(MASS)
predicted_transa_robust <- harju %>% 
  rlm(transactions ~ month + area, data = ., weights = price_min) %>% 
  augment
```

Add robust fit to plot:
```{r}
p + geom_line(data = predicted_transa, aes(month, .fitted, group = 1), 
              color = "red",
              size = 1) + 
  geom_line(data = predicted_transa_robust, aes(month, .fitted, group = 1), 
            color = "blue",
            size = 1) +
  labs(caption = "Red line: model fit of seasonal effect to number of transactions.\nBlue line: robust model fit of seasonal effect to number of transactions, weighted for minimal price. \nSource: Estonian Land Board, transactions database.")
```
## Whole dataset: all counties

Plot whole dataset:
```{r}
apartments %>% 
  ggplot(aes(date, transactions, group = area, color = area)) +
  geom_line() +
  facet_wrap(~ county, scales = "free_y")
```


```{r}
apartments %>% 
  mutate(price_unit_area_mean = price_unit_area_mean / consumerindex) %>% 
  ggplot(aes(date, price_unit_area_mean, group = area, color = area)) +
  geom_line() +
  facet_wrap(~ county, scales = "free_y")
```


## Fit multiple models

Fit model for each county. 
Goodness of model fit is evaluated by its ability to predict response values.
Response values can be predicted using the same data that was used to train model.
In this case we are dealing with in-sample analysis, which is ok when all we want to know is model coeficients (eg. intercept and slope) to describe present data.
In cases when we want to predict using some unseen future data, such models are usually overfitted, overly optimistic and don't perform very well.
To manage overfitting, we split dataset into training set for model fitting and test set for model performance assessment.

Resampling can be performed very well using base R function `sample()`.
Let's say we want to use approx. 70% of rows from "mtcars" table to be used in model training and leave remaining 30% for model testing.
Basically, all we need to do is to generate to non-overlapping row indexes.
```{r}
set.seed(123)
index <- sample(1:nrow(mtcars), floor(0.7 * nrow(mtcars)), replace = FALSE)
index
## ~70% data goes to training set
train <- mtcars[index, ]
## remaining ~30% will be used for testing model performance
test <- mtcars[-index, ]
```

Index based resampling is also implemented in tidyverse.
We can use `resample_partition()` function from "modelr" library to perform.
Here we generate data splits and move them to separate columns.
```{r}
library(modelr)
## Generate data splits and move into separate columns
trans <- apartments %>%
  group_by(county) %>%
  nest %>% 
  mutate(parts = map(data, ~resample_partition(.x, c(test = 0.3, train = 0.7))),
         test = map(parts, "test"),
         train = map(parts, "train")) %>% 
  dplyr::select(-parts)
```

Fit model to each data split. 
Let's start with number of transactions.
```{r}
## Fit model
trans <- trans %>%
  mutate(mod_train = map(train, ~lm(transactions ~ month + area, data = .x)))
## Print data with new model column
trans %>% 
  dplyr::select(mod_train, everything())
```



`coef()` and `summary()` return vector and S3 object respectively. 
Model coeficients and summary statistics can be returned as a data frame using `tidy()` function from "broom" library:
```{r}
## model coeficients
trans %>% 
  mutate(mod_tidy = map(mod_train, tidy)) %>% 
  dplyr::select(county, mod_tidy) %>% 
  unnest %>% 
  head()
```


Pseudo out-of-sample model performance, `augment()` adds predictions to original data frame (.fitted).
We subtract fitted values from the original values and calculate root-mean-squared deviation (rmsd).
Smaller rmsd indicates better fit.
```{r}
trans <- trans %>%
  mutate(mod_pred = map2(mod_train, test, ~augment(.x, newdata = .y)),
         mod_pred = map(mod_pred, ~mutate(.x, .resid = transactions - .fitted)),
         mod_rmsd = map_dbl(mod_pred, ~sqrt(mean(.x$`.resid`^2)))) 
trans %>% 
  dplyr::select(county, mod_rmsd) %>% 
  arrange(desc(mod_rmsd))
```
We can see that Harju county displays worst fit and Tartu is next. These two counties show also majority of transactions. 

```{r}
trans_pred <- trans %>% 
  dplyr::select(county, mod_pred) %>% 
  unnest
trans_pred %>% 
  dplyr::select(.fitted, transactions, everything()) %>% 
  head()
```

