---
title: "Modelling"
author: "Taavi Päll"
date: "17 10 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Estonian Land Board data

Available via [www.maaamet.ee](http://www.maaamet.ee/kinnisvara/htraru/FilterUI.aspx).
API can be accessed programmatically using html requests.
One can use "rvest" package. Please have a look at "R/maaamet.R" script in [rstats-tartu/datasets](https://www.github.com/rstats-tartu/datasets) repo.
Returns html, but that ok too.
Html results need little wrangling to get table.

Data that can be downloaded are number of transactions and price summaries per property size, and type for different time periods.
Price info is given for data splits with more than 5 transactions.

## Load dataset

```{r}
##  Check if file is alredy downloaded 
if(!file.exists("data/transactions_residential_apartments.csv")){
  url <- "https://raw.githubusercontent.com/rstats-tartu/datasets/master/transactions_residential_apartments.csv"
  
  ## Check if data folder is present
  if(!dir.exists("data")){
    dir.create("data")
  }
  ## Download file to data folder
  download.file(url, "data/transactions_residential_apartments.csv")
}
```

## Import
```{r}
library(tidyverse)
library(lubridate)
library(stringr)
library(viridis)
apartments <- read_csv("data/transactions_residential_apartments.csv")
apartments <- apartments %>% 
  mutate(date = ymd(str_c(year, month, 1, sep = "-"))) %>% 
  select(date, everything())
harju <- apartments %>% filter(str_detect(county, "Harju"))
harju
```

## Strategy

- Start with single unit and identify interesting pattern

- Summarise pattern with model

- Apply model to all units

- Look for units that don't fit pattern

- Summarise with single model


## First glimpse

Plot **number of transctions** per year for Harju county and add smooth line.
```{r}
p <- ggplot(harju, aes(factor(year), transactions, group = area, color = area)) +
  geom_point() +
  geom_smooth(method = 'loess') +
  facet_wrap(~county, scales = "free_y") +
  labs(
    y = "Transactions",
    x = "Year"
  ) +
  scale_color_viridis(discrete = TRUE)
p
```

Mean price per unit area:
```{r}
## Add date 
p <- harju %>%
  ggplot(aes(date, price_unit_area_mean, group = area, color = area)) +
  geom_line() +
  geom_vline(xintercept = ymd("2008-09-15"), linetype = 2) +
  labs(title = "Transactions with residential apartments",
       subtitle = "Harju county",
       x = "Date",
       y = bquote(Mean~price~(eur/m^2)),
       caption = str_c("Source: Estonian Land Board, transactions database.\nDashed line, the collapse of the investment bank\nLehman Brothers on Sep 15, 2008.")) +
  scale_color_viridis(discrete = TRUE, name = bquote(Apartment~size~m^2))
p
```

Adjust mean price to changes in consumer index:
```{r}
download.file("https://raw.githubusercontent.com/rstats-tartu/datasets/master/consumer_index.csv",
              "data/consumer_index.csv")
consumer_index <- read_csv("data/consumer_index.csv")
consumer_index
## check if 2005 is 100%
consumer_index[10, 3:14] %>% rowMeans()
divide_by_100 <- function(x) {
  x/100
  }
consumer_index <- consumer_index %>% 
  select(-X1) %>% 
  gather(key = month, value = consumerindex, -year) %>% 
  mutate_at("consumerindex", divide_by_100)
consumer_index
```

Join apartments and consumer index data:
```{r}
apartments <- left_join(apartments, consumer_index, by = c("year", "month"))
```

Select harju county data:
```{r}
harju_ci <- filter(apartments, str_detect(county, "Harju"))
harju_ci
harju_ci <- harju_ci %>% 
  mutate(pua_mean_ci = price_unit_area_mean / consumerindex)
```


Price per m^2 after consumer index adjustment:
```{r}
harju_ci %>%
  ggplot(aes(date, pua_mean_ci, group = area, color = area)) +
  geom_line() +
  geom_vline(xintercept = ymd("2008-09-15"), linetype = 2) +
  labs(title = "Transactions with residential apartments",
       subtitle = "Harju county",
       x = "Date",
       y = "Consumer index corrected mean price\nrelative to 2005 (eur/m^2)",
       caption = str_c("Source: Estonian Land Board, transactions database.\nDashed line, the collapse of the investment bank\nLehman Brothers on Sep 15, 2008.")) +
  scale_color_viridis(discrete = TRUE, name = bquote(Apartment~size~m^2))
```

Seasonal pattern? Mean price eur/m^2 per month:
```{r}
harju_per_month <- harju %>% 
  group_by(year, month, area) %>% 
  summarise_at(vars(transactions,
                    area_mean,
                    price_unit_area_mean), 
               mean, na.rm = TRUE)
harju_per_month # factor(month, levels = month.abb)
p <- harju_per_month %>%
  ggplot(aes(factor(month, levels = month.abb), transactions, group = year)) +
  geom_line(alpha = 0.3) +
  geom_vline(xintercept = ymd("2008-09-15"), linetype = 2) +
  facet_wrap(~ area) +
  labs(title = "Transactions with residential apartments",
       subtitle = "Harju county",
       x = "Month",
       y = "Transactions",
       caption = str_c("Source: Estonian Land Board, transactions database.\nDashed line, the collapse of the investment bank\nLehman Brothers on Sep 15, 2008.")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
p
```

Add trendline per month:
```{r}
# install.packages("broom")
library(broom)
predicted_hpm <- harju_per_month %>% 
  lm(transactions ~ month + area, data = .) %>% 
  fortify()
predicted_hpm
p + geom_line(data = predicted_hpm, aes(month, .fitted, group = 1), 
              color = "red", 
              size = 2)
```
Small trend at end of year?


Number of transactions per month:
```{r}
p <- harju_per_month %>%
  ggplot(aes(factor(month, levels = month.abb), transactions, group = year)) +
  geom_line(alpha = 0.3) +
  geom_vline(xintercept = ymd("2008-09-15"), linetype = 2) +
  facet_wrap(~ area) +
  labs(title = "Transactions with residential apartments",
       subtitle = "Harju county",
       x = "Month",
       y = "Number of transcations",
       caption = str_c("Source: Estonian Land Board, transactions database.\nDashed line, the collapse of the investment bank\nLehman Brothers on Sep 15, 2008.")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
predicted_hpm <- harju_per_month %>% 
  lm(transactions ~ month + area, data = .) %>% 
  fortify()
p + geom_line(data = predicted_hpm, aes(month, .fitted, group = 1), 
              color = "red", 
              size = 1)
```

Try to remove seasonal pattern and look then at the longer trend:
```{r}

```



